### 2015-04-22 doc/thesis_puc/abstract.tex
Adding abstract and appendix

Dynamically typed languages such as Lua avoid static types in favor of
simplicity and flexibility, because the absence of static types means
that programmers do not need to bother with abstracting types that
should be validated by a type checker.
In contrast, statically typed languages provide the early detection of
many bugs, and a better framework for structuring large programs.
These are two advantages of static typing that may lead programmers
to migrate from a dynamically typed to a statically typed language,
when their simple scripts evolve into complex programs.
Optional type systems allow combining dynamic and static typing in
the same language, without influencing its original semantics,
making easier this code evolution from dynamic to static typing.
Designing an optional type system for a dynamically typed language
is challenging, as it should feels natural to programmers that are
already familiar with this language.
In this work we present and formalize the design of Typed Lua,
an optional type system for Lua that introduces novel features
to statically type check some Lua idioms and features.
Even though Lua shares several characteristics with other dynamically
typed languages such as JavaScript, Lua also has several unusual features
that are not present in the type system of these languages.
We discuss how Typed Lua handles features such as functions with flexible
arity that interact with multiple assignment, functions that are overloaded
on the number of return values, and the incremental evolution of record and object types.
We also discuss our design decisions, and present the evaluation results
that we achieved while using Typed Lua to type existing Lua code.

>>>>

Dynamically typed languages such as Lua avoid static types in favor of
simplicity and flexibility, because the absence of static types means
that programmers do not need to bother with abstracting types that
should be validated by a type checker.
In contrast, statically typed languages provide the early detection of
many bugs, and a better framework for structuring large programs.
These are two advantages of static typing that may lead programmers
to migrate from a dynamically typed to a statically typed language,
when their simple scripts evolve into complex programs.

Optional type systems allow combining dynamic and static typing in
the same language, without influencing its original semantics,
making easier this code evolution from dynamic to static typing.
Designing an optional type system for a dynamically typed language
is challenging, as it should feels natural to programmers that are
already familiar with this language.

In this work we present and formalize the design of Typed Lua,
an optional type system for Lua that introduces novel features
to statically type check some Lua idioms and features.
Even though Lua shares several characteristics with other dynamically
typed languages such as JavaScript, Lua also has several unusual features
that are not present in the type system of these languages.
These features include functions with flexible arity, multiple assignment,
functions that are overloaded on the number of return values, and the
incremental evolution of record and object types.
We discuss how Typed Lua handles these features and our design decisions.
Finally, we present the evaluation results that we achieved while using
Typed Lua to type existing Lua code.

### 2015-04-22 doc/thesis_puc/resumo.tex
Adding abstract and appendix

Linguagens dinamicamente tipadas, tais como Lua, não usam tipos estáticos em
favor de simplicidade e flexibilidade, porque a ausência de tipos estáticos
significa que programadores não precisam se preocupar em abstrair tipos que
devem ser validados por um verificador de tipos.
Por outro lado, linguagens estaticamente tipadas ajudam na detecção prévia de
diversos \emph{bugs} e também ajudam na estruturação de programas grandes.
Tais pontos geralmente são vistos como duas vantagens que levam programadores
a migrar de uma linguagem dinamicamente tipada para uma linguagem estaticamente tipada,
quando os pequenos \emph{scripts} deles evoluem para programas complexos.
Sistemas de tipos opcionais nos permitem combinar tipagem dinâmica e estática na
mesma linguagem, sem influenciar a semântica original da linguagem, tornando mais
fácil a evolução de código tipado dinamicamente para código tipado estaticamente.
Desenvolver um sistema de tipos opcional para uma linguagem dinamicamente tipada é
uma tarefa desafiadora, pois ele deve ser o mais natural possível para os programadores
que já estão familiarizados com essa linguagem.
Neste trabalho nós apresentamos e formalizamos Typed Lua, um sistema de tipos opcional
para Lua, o qual introduz novas características para tipar estaticamente alguns idiomas
e características de Lua.
Embora Lua compartilha várias características com outras linguagens dinamicamente
tipadas, em particular JavaScript, Lua também possui várias características não usuais,
as quais não estão presentes nos sistemas de tipos dessas linguagens.
Nós discutimos como Typed Lua tipa estaticamente características tais como
funções com aridade flexível que interagem com atribuições múltiplas,
funções que são sobrecarregadas no número de valores de retorno e a evolução
incremental de estruturas e objetos.
Discutimos também nossas decisões de projeto e apresentamos uma avaliação de
resultados, a qual obtemos ao usar Typed Lua para tipar código Lua existente.

>>>>

Linguagens dinamicamente tipadas, tais como Lua, não usam tipos estáticos em
favor de simplicidade e flexibilidade, porque a ausência de tipos estáticos
significa que programadores não precisam se preocupar em abstrair tipos que
devem ser validados por um verificador de tipos.
Por outro lado, linguagens estaticamente tipadas ajudam na detecção prévia de
diversos \emph{bugs} e também ajudam na estruturação de programas grandes.
Tais pontos geralmente são vistos como duas vantagens que levam programadores
a migrar de uma linguagem dinamicamente tipada para uma linguagem estaticamente tipada,
quando os pequenos \emph{scripts} deles evoluem para programas complexos.

Sistemas de tipos opcionais nos permitem combinar tipagem dinâmica e estática na
mesma linguagem, sem influenciar a semântica original da linguagem, tornando mais
fácil a evolução de código tipado dinamicamente para código tipado estaticamente.
Desenvolver um sistema de tipos opcional para uma linguagem dinamicamente tipada é
uma tarefa desafiadora, pois ele deve ser o mais natural possível para os programadores
que já estão familiarizados com essa linguagem.

Neste trabalho nós apresentamos e formalizamos Typed Lua, um sistema de tipos opcional
para Lua, o qual introduz novas características para tipar estaticamente alguns idiomas
e características de Lua.
Embora Lua compartilhe várias características com outras linguagens dinamicamente
tipadas, em particular JavaScript, Lua também possui várias características não usuais,
as quais não estão presentes nos sistemas de tipos dessas linguagens.
Essas características incluem funções com aridade flexível, atribuições múltiplas,
funções que são sobrecarregadas no número de valores de retorno e
a evolução incremental de registros e objetos.
Nós discutimos como Typed Lua tipa estaticamente essas características e
também discutimos nossas decisões de projeto.
Finalmente, apresentamos uma avaliação de resultados,
a qual obtemos ao usar Typed Lua para tipar código Lua existente.
